---
title: "STA 207 Final - Inference"
format: html
---

# Data loading and Libraries

```{r}
load("C:/Users/user/Documents/GitHub/STA207/sta207-project/data/df_subset.rda")

# Convert column "subject" into factor
df_subset$subject <- as.factor(df_subset$subject)

# Libraries
library(dplyr)
library(car)
library(lme4)
library(influence.ME)
```

<!-- 
# Check For Collinearity

For each school, it has a unique `urban` status.
```{r}
# Shows a school that has the most urban type(s)
check_nesting <- df_subset %>%
  group_by(schid) %>%
  summarize(n_urban_types = n_distinct(urban)) %>%
  summarize(max_n_urban = max(n_urban_types))

#print(check_nesting)
```

All students go to one school only.
```{r}
# Shows a school that has the most urban type(s)
check_stdnt_sch <- df_subset %>%
  group_by(stdntid) %>%
  summarize(n_schs = n_distinct(schid)) %>%
  summarize(max_n_schs = max(n_schs))

#print(check_stdnt_sch)
```
-->

# Inferential Analysis

## Model Proposal:

### Multiple Linear Regression

To find the best multiple linear model, we begin with our full model selected from our EDA.
We make the following assumptions for our models:

1. Linearity: The relationship between predictors and the outcome is linear.
2. Independence: Observations (students) are independent.
3. Homoscedasticity: The variance of error terms is constant across groups.
4. Normality: The residuals are approximately normally distributed.

```{r}
full_model <- lm(score ~ class_type+subject+gender+race+grade+urban, df_subset)
summary(full_model)
```

From the summary statistic, we see that all the variables are signifcantly associated with score.

#### Include Interaction Terms

Including the interaction of subject and grade significantly reduced the curvature of the residuals versus fitted values plot, so we include this interaction term for the following models.
```{r}
plot(full_model, which=1)

inter_full <- lm(score ~ class_type+subject+gender+grade+urban+race+subject*grade, df_subset)
summary(inter_full)
plot(inter_full, which=1)
```

#### ANOVA With Type III Sum of Square

We consider running ANOVA on this full model using Type III Sum of Squares:
```{r}
Anova(inter_full, type="III")
```

From the summaries of multiple linear model and the ANOVA, we don't find any variable that is obviously irrelavant to scores. Thus, we keep all these variables and continue to fit our mixed effect model to discover causal relationship.

### Mixed Models
We make the following assumptions for our mixed models:

1. Linearity: The relationship between predictors and the outcome is linear.
2. Independence: Observations (students) are independent.
3. Homoscedasticity: The variance of error terms is constant across groups.
4. Normality: The residuals are approximately normally distributed.
Stratifying using `schid`: Grouping students by schools.

```{r}
mixed_model_sch <- lmer(score
    ~ class_type
    + subject
    + gender
    + race
    + grade
    + subject*grade
    + (1|schid),
    data=df_subset)

summary(mixed_model_sch)
```

Stratifying using `stdntid`: Consider each student as a group.
```{r}
mixed_model_std <- lmer(score
    ~ class_type
    + subject
    + gender
    + race
    + grade
    + subject*grade
    + (1|stdntid),
    data=df_subset)

summary(mixed_model_std)
```

Stratifying using both `schid` and `stdntid`: See how much does school affect the score.
```{r}
mixed_model_std_sch <- lmer(score
    ~ class_type
    + subject
    + gender
    + race
    + grade
    + subject*grade
    + (1|stdntid)
    + (1|schid),
    data=df_subset)

summary(mixed_model_std_sch)
```
<!--
```{r}
mixed_model_std_sch_grade <- lmer(score
    ~ class_type
    + subject
    + gender
    + race
    + subject*grade
    + (grade|stdntid),
    data=df_subset)

anova(mixed_model_std, mixed_model_std_sch_grade)
```
-->

# Sensitivity Analysis

```{r}
# Get a list of all unique school IDs
all_schools <- unique(df_subset$schid)
# Create a vector to store results
results <- numeric(length(all_schools))

#for(i in 1:length(all_schools)) {
for(i in 1:3) {
  # Excluding one school
  temp_data <- subset(df_subset, schid != all_schools[i])
  
  #temp_model <- lmer(score ~ class_type + grade + gender + subject + race + subject*grade + (1|stdntid), data = temp_data)
  temp_model <- lmer(score ~ class_type + grade + gender + subject + race + subject*grade + (1|stdntid), data = temp_data)
  
  # Coefficient for small class size
  results[i] <- fixef(temp_model)["class_typesmall"]
}

plot(results, main="Sensitivity: Leave-One-School-Out", ylab="Small Class Coefficient")
abline(h = mean(results), col="red")
```

```{r}
# Calculate influence of schools
# influence_school <- influence(mixed_model_std_sch, group="schid")

# Plot Cook's Distance
# plot(influence_school, which="cook")
```

# Plausibility Check
<!-- TO-DO: Finish plausibility check for the final model -->
## Linearity Check

```{r}
plot(mixed_model_std_sch)
```

## Normality of Residuals and Random Effects
```{r}
# Plot the histogram of residuals
res <- residuals(mixed_model_std_sch)

# Plot the histogram
hist(res, 
     breaks = 50, 
     col = "skyblue", 
     xlim = c(-100, 100),
     main = "Histogram of Residuals", 
     xlab = "Residuals")

x <- seq(min(res), max(res), length = 100)
y <- dnorm(x, mean = mean(res), sd = sd(res))/3.6
y <- y * diff(hist(res, plot = FALSE)$mids[1:2]) * length(res)
lines(x, y, col = "red", lwd = 2)
```

```{r}
# Plot the histogram for random effect: student
random_effects <- ranef(mixed_model_std_sch)
student_intercepts <- random_effects$stdntid[, "(Intercept)"]
hist(student_intercepts, 
     breaks = 20, 
     col = "darkseagreen", 
     main = "Histogram of Random Intercepts (Students)", 
     xlab = "Random Effect Estimate")
```

```{r}
# Plot the histogram for random effect: school
# Since we have 75 schools only, the normality isn't visible
random_effects <- ranef(mixed_model_std_sch)
school_intercepts <- random_effects$schid[, "(Intercept)"]
hist(student_intercepts, 
     breaks = 7, 
     col = "darkseagreen", 
     main = "Histogram of Random Intercepts (Schools)", 
     xlab = "Random Effect Estimate")
```

## Homoscedasticity
```{r}
plot(mixed_model_std_sch)
```

<!--
## Independence
```{r}
acf(res, main="ACF of Model Residuals")
```
-->

# Conclusion

Small class indeed leads to better students' performance, regardless of which school they attend or the innate intellegence each student have.


<!-- TO-DO: Create images of the models' equations -->